---
title: "Project_01"
output: html_document
date: "2024-02-16"
---
### PART 1
```{r}
# Part 1
 
#This just loading all packages used in hw 

library(dplyr)

library(tibble)

library(lubridate)

library(readr)

library(ggplot2)
```


#### Create a dataframe with 4 variables.
```{r}
#Make the dataframe literally as mentioned in instructions

doctor_data <- data.frame(

  doctor_type = c("PCP", "Psychiatrist", "Surgeon", "Anesthesia"),

  doctor_lastname = c("Smith", "Dame", "Jones", "Zayas"),

  location = c("MA", "ME", "NH", "VT"),

  AVG_Rating = c(7, 9, 8, 9)

)
 
# Print the thing as mentioend by instructions

print(doctor_data)
```


#### Select row 1 in column 2, what was selected?
#### Select rows 2 through 4, what was selected?
#### Select the last column, what was selected?
```{r}
selected_1_2 <- doctor_data[1, 2]  # Take the 1 row and 2 column. I got Smith
selected_1_2 

selected_2_to_4 <- doctor_data[2:4, ]  #  Take the rows 2 through 4 for all columns
selected_2_to_4

selected_last_column <- doctor_data[, 4]  # Pick all rows in the fourth column
selected_last_column 

```


####  Using the dataset Mtcars create a scatterplot.
```{r}
# Get column names
colnames(mtcars)
 
# Scatterplot

ggplot(mtcars, aes(x = mpg, y = hp, color = as.factor(cyl))) +  # Convert cyl to factor for discrete color scale

  geom_point() +

  labs(title = "MPG vs Horsepower in mtcars Dataset",

       x = "Miles Per Gallon (MPG)",

       y = "Horsepower (HP)",

       color = "Number of Cylinders") +

  theme_minimal()
 
## Plot shows the relationship between HP(Horsepower) and MPG(Miles per Gallon)

## of various cars in dataset. What I observe/see is an inverse relationship: cars

## with higher horsepower tend to have lower Miles Per Gallon. Yet again, I know

## very little about cars but I thought higher horspower means higher miles per

## gallon so interesting to see the opposite. Also thats why i picked the two

## variables.

## The graph also shows the data for cylinders The trend is that cars with more

## cylinders have lower mpg
```


####  Using the dataset MTcars perform a summary statistic on the dataset and calculate the pearson coefficient.
```{r}
# Summary Statistics 

summary(mtcars)
 
# TAKE Pearson Correlation Coefficient between 'wt' and 'mpg'

correlation_coefficient <- cor(mtcars$wt, mtcars$mpg, method = "pearson")
 
# Print coeffiecnt

print(correlation_coefficient)
 
# The Pearson corrleation coefficent for 'wt' and 'mpg' was calculated to be -0.867

# This statistic is a stat measure of linear relationship between two variabls

#A value of -0.867 indicates a strong negative linear relationship between the weight of the cars and their fuel efficiency. 

#As the weight of the car increases, the fuel efficiency (measured in miles per gallon) tends to decrease. 

#This result is heavier cars generally need more energy to move which causes lower fuel efficiency. NOT A CAR EXPERT BTW.
```


### PART2

####  Load the data into your R environment.
```{r}

# This just loads data
data <- read_csv("Substance_Use_Disorder_Treatment_Program_Admissions__Beginning_2007_20240202.csv")

head(data)
```


####  Evaluate the dataset, removed/imputed outliers, perform the correct summary statistics. Provided the correct justification.
```{r}
dim(data)
# 99,367 rows and 7 columns
```


```{r}
glimpse(data)
# This just a glimpse of data
```


```{r}
summary(data)
# This shows various statistics available
```


```{r}
# Interesting to see Admissions
# Min of 1, 1st quartile 2, median 8, Mean 41.91, 3rd quartile 28, Max 2861
# Will graph to see skewsness, I think this is right skewed.
```


```{r}
# Basic statistics for the 'Admissions' column
admissions_desc <- summary(data$Admissions)
admissions_desc

# Plotting histogram using base R
hist(data$Admissions, 
     main = "Histogram of Admissions", 
     xlab = "Admissions", 
     col = "skyblue", 
     border = "white")
 
# Skewed to the right but graph is not the best
```


```{r}
# NA
missing_values <- sum(is.na(data))
missing_values
 
# 0 missing values
```


```{r}
frequency_table <- table(data$`Primary Substance Group`)
frequency_table
```


```{r}
# This is me filtering out the rows where 'Primary Substance Group' is 'None'
data <- subset(data, `Primary Substance Group` != "None")

# We did this to remove the "None" value because there was only one observation which contained "None" as there Primary Substance Group which can be considered as wrong data value.
table(data$`Primary Substance Group`)
```


#### In normal distribution, most data falls within 3 standard deviations of the mean. Basically from stats class any Z-scores above 3 or below -3 could basically mean outliers. Then values with Z-scores beyond 3 are treated as outliers and replaced with the median as we did.
```{r}
# Calculating Z- score
data$z_score <- abs(mean(data$Admissions) - data$Admissions) / sd(data$Admissions)


#Imputing all values which have z score more than 3 with the median of the values
data[data$z_score > 3, "Admissions" ] <- median(data$Admissions)

```


#### four tibbles or data frames - 
```{r}
# We are just writing out the vectors for county codes and names manually because it did not work otherwise
county_codes <- c("AL", "AG", "BX", "BM", "CA", "CY", "CH", "CM", "CN", "CL", "CO", "CR", "DE", "DU", "ER", "ES", "FR", "FU", "GE", "GR", "HE", "JE", "KG", "LE", "LI", "MA", "MO", "MG", "NA", "NY", "NI", "ON", "OD", "OT", "OR", "OL", "OS", "OG", "PU", "QN", "RE", "RD", "RO", "SL", "SA", "SC", "SH", "SY", "SE", "ST", "SU", "SV", "TI", "TO", "UL", "WR", "WS", "WA", "WE", "WY", "YA")

county_names <- c("Albany", "Allegany", "Bronx", "Broome", "Cattaraugus", "Cayuga", "Chautauqua", "Chemung", "Chenango", "Clinton", "Columbia", "Cortland", "Delaware", "Dutchess", "Erie", "Essex", "Franklin", "Fulton", "Genesee", "Greene", "Herkimer", "Jefferson", "Kings", "Lewis", "Livingston", "Madison", "Monroe", "Montgomery", "Nassau", "New_York_City", "Niagara", "Oneida", "Onondaga", "Ontario", "Orange", "Orleans", "Oswego", "Otsego", "Putnam", "Queens", "Rensselaer", "Richmond", "Rockland", "St_Lawrence", "Saratoga", "Schenectady", "Schoharie", "Schuyler", "Seneca", "Steuben", "Suffolk", "Sullivan", "Tioga", "Tompkins", "Ulster", "Warren", "Washington", "Wayne", "Westchester", "Wyoming", "Yates")
 
# Tibble
county_tibble <- tibble(county_code = county_codes, county_name = county_names)
 
# Print the tibble to check
print(county_tibble)
 

```



```{r}
# Create program codes
program_codes <- c("Crisis" = "CR", "Inpatient" = "IN", "Opioid Treatment Program" = "OT", 
                   "Outpatient" = "OP", "Residential" = "RE", "Specialized" = "SP")
 
# This is me literally asssigning  the program codes to the Program.Category
program_category_df <- data %>%
  distinct(`Program Category`) %>%
  mutate(program_code = program_codes[`Program Category`])
 
# Check
print(program_category_df)
```


```{r}
# Create substance codes
substance_codes <- c("Heroin" = "H", "All Others" = "AO", "Other Opioids" = "OO", 
                     "Alcohol" = "A", "Cocaine" = "C", "Marijuana" = "M")
 
# Just asssigning the substance codes to the Primary.Substance.Group 
primary_substance_group_df <- data %>%
  distinct(`Primary Substance Group`) %>%
  mutate(substance_code = substance_codes[`Primary Substance Group`])

# Check
print(primary_substance_group_df)
```


```{r}
 # Now last one is the biggest one
admissions_data <- data %>%
  left_join (county_tibble, by = c("County of Program Location" = "county_name")) %>%
  left_join(program_category_df, by = c("Program Category" = "Program Category")) %>%
  left_join(primary_substance_group_df, by = c("Primary Substance Group" = "Primary Substance Group")) %>%
  select(Year, 
          'County of Program Location' = county_code,  # county_tibble
          'Program Category' = program_code,           # program category
         `Service Type`, 
         `Age Group`, 
         `Primary Substance Group` = substance_code,  # primary_substance_group_df
         Admissions)
 
# first few
head(admissions_data)
```


#### Create a function called annualAdmissions() 
```{r}
annualAdmissions <- function(data) {
  # Summarize the total admissions by year
  admissions_by_year <- data %>%
    group_by(Year) %>%
    summarise(Total_Adm = sum(Admissions, na.rm = TRUE)) %>%
    arrange(desc(Total_Adm))
 
  # Finding  the year with the highest number of admissions
  top_year <- admissions_by_year %>%
    filter(Total_Adm == max(Total_Adm))
 
  # Plot
  p <- ggplot(admissions_by_year, aes(x = Year, y = Total_Adm)) +
    geom_line() +
    geom_point() +
    annotate("text", x = top_year$Year, y = max(admissions_by_year$Total_Adm),
             label = paste("Max Year:", top_year$Year),
             vjust = -0.5, hjust = 1, color = "red") +  # Added annotate function
    scale_x_continuous(breaks = unique(admissions_by_year$Year)) +  # Set x breaks to every year
    labs(title = "Total Number of Admissions per Year in NY",
         x = "Year",
         y = "Number of Admissions") +
    theme_minimal()
  
  print(p)
  
  # Explanation
  cat("The chart shows the total number of admissions per year for the entire state of New York",
      "Each year is displayed on the x-axis. The year with the highest number of admissions is annotated basically shown",
      "It shows trends over time.")
}

annualAdmissions(admissions_data)
```

#### Analyze the percentage of admissions for each county and visualize the results for the top 5 counties using a bar chart. Explain the results.
```{r}
# We calculate the total admissions for each county
total_admissions_by_county <- data %>%
  
  group_by(`County of Program Location`) %>%
  
  summarise(Total_Adm = sum(Admissions, na.rm = TRUE)) %>%
  
  ungroup()

total_admissions_by_county
```


```{r}
# We are using the total number of admissions to find percentages
grand_total_admissions <- sum(total_admissions_by_county$Total_Adm)
 
# We are using the % of total admissions for each county
total_admissions_by_county <- total_admissions_by_county %>%
 
   mutate(Percentage = (Total_Adm / grand_total_admissions) * 100)


total_admissions_by_county[ 1:5, c("County of Program Location" , "Percentage")]

```


```{r}
# Geting the top 5 counties by percentage
top_5_counties <- total_admissions_by_county %>%
  
  arrange(desc(Percentage)) %>%
  
  slice_head(n = 5)

top_5_counties
```


```{r}
# Plotting
top_5_counties %>%

ggplot(aes(`County of Program Location` , Percentage)) + 
  
  geom_bar(stat = "identity", fill = 'skyblue') +
  
  geom_text(aes(label = paste(round(Percentage, 2), "%")), vjust = -0.5) +
  
  labs(title = "Top 5 Counties by Percentage of Admissions",
       
       x = "County",
       y = "Percentage of Total Admissions") +
  
  scale_fill_brewer(palette = "Set3") +  # Use a categorical color palette
  
  theme_minimal()
 
## Erie, Kings, Queens, Suffolk and New York are presesnted in the graph



```

#### Filter the data, using a regular expression, and extract all admissions to the various “Rehab” facilities
```{r}
# Creating dataframe called Rehab_data which filters all facilities that include the word rehab, rehabilitation
Rehab_data <- data %>%
     
  filter(grepl("rehab|rehabilitation", `Service Type`, ignore.case = TRUE))
 
head(Rehab_data,3)
```


#### Using the filtered data, identify which substance is the most prominent among each age group.
```{r}
rehab_data <- Rehab_data %>% 
  
  select(`Age Group`, `Primary Substance Group`, Admissions) %>%
  
  left_join(primary_substance_group_df, by = c("Primary Substance Group" = "substance_code")) %>%
  
  group_by(`Age Group`, `Primary Substance Group`) %>%
  
  summarise(total_admissions = sum(Admissions))


head(rehab_data , 3)
```


```{r}
rehab_data %>%
  
  group_by(`Age Group`) %>% 
  
  top_n(1, total_admissions) 
```


```{r}
rehab_data %>%
  
  group_by(`Age Group`) %>% 
  
  top_n(1, total_admissions) %>%

ggplot(aes(`Age Group`, total_admissions, fill = `Primary Substance Group`)) +
  
  geom_bar(stat = "identity", position = "dodge") +
  
 
  
  labs(title = "Most Prominent Substance in Each Age Group for 'Rehab' Facilities",
       x = "Most Prominent Substance",
       y = "Age Group") +
  
  theme_minimal() +
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

